Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-04-01 index.qmd
3/30 update
Update 3/31

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(ezcox)
library(sjPlot)

@@

plot_model(forest,
           dot.size = 1,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
show_forest(lung, covariates = c('age'))

@@

### Prediction

@@

Though CPH is typically used with health data, we wanted to determine
how CPH can be used in other fields. As such, we will study the
remaining useful life (RUL) of a NASA aircraft engine using data
provided by NASA for the study on propagation modeling [@saxena2008]. In
particular, we will be examining their data from what they have deemed
engine two. From the literature review, we have noted that we will need
to be concerned with the constant hazards ratio within our data and
determine how we will accommodate censored data when creating our model.

#### Data Structure

Each engine has an unknown amount of wear, manufacturing variation, and
sensor noise all of which impact the survival time. In the training
dataset, each engine is operating normally at the beginning of the time
series, develops a fault, and experiences failure at the end of the time
series. The conditions the engines are subjected to are represented by
three operational setting fields and 21 sensor measurement fields. In
the testing data, the engines have not experienced failure
[@saxena2008].

For both test and train datasets, we will need to add a column
indicating status. We will use 0 for still working and 1 for the last
cycle when the engine fails. For the test data, we know that none of the
engines have failed so all values in the test column will be 0 for still
working. For the training data, we know that the data includes all
cycles up to the last cycle when the machine will fail. As such, we will
have 0s in all cycles except the last cycle of the machine.

@@

**Testing Data**

@@

kable(head(test, n = 5), format = "markdown")
```

**Training Data**

```{r}
trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'

@@

kable(head(train, n = 5), format = "markdown")
```

#### Data Visualization

The visuals below describe the distribution of the 260 engines in the
training data. The histogram shows the survival times are normally
distributed with a median value of 199 iterations and the box plot shows
there are few outliers out of the 260 engines. The engines have a median
survival time of 199 iterations with a standard deviation of 46.8. The
longest surviving engine lasted 378 iterations, with the shortest
survival time for all engines being 128.

```{r}
distributions <- train %>% 
  group_by(id) %>%
  summarise(time = max(time))

@@

colnames(summarymetrics) <- c("Metric", "Value")
summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
kable(summarymetrics)

@@

The Kaplan-Meier survival curve is shown below.

@@

#### Choosing Covariates

Below is a univariate test to determine which covariates to use in the
model.

@@

#### Creating Model

To generate the CPH model with properly formatted survival data, a
survival object must first be created using the R package `survival`.
The survival object is created by running `Surv(time, event)` where
`time` is the time field and `event` is the event field.

To create the CPH model, the survival object is then applied to the R
function `coxph(survival object ~ x, data)` where `x` represents one or
multiple covariate(s) and `data` is the survival data that is being
modeled.[@R-base].

During the analysis, multiple models will be created with a different
combination of covariates.

@@

The resulting models and their corresponding concordance index values
are below.

```{r}
concordance_results <- data.frame(row.names = c(
  "Model 1",
  "Model 2",
  "Model 3",
  "Model 4",
  "Model 5",
  "Model 6",
  "Model 7",
  "Model 8",
  "Model 9",
  "Model 10"),
  Concordance = c(
  cox1$concordance[6],
  cox2$concordance[6],
  cox3$concordance[6],
  cox4$concordance[6],
  cox5$concordance[6],
  cox6$concordance[6],
  cox7$concordance[6],
  cox8$concordance[6],
  cox9$concordance[6],
  cox10$concordance[6]))

kable(concordance_results)

@@

The analysis will proceed using model 1 because of the high concordance
index. The full summary of model 1 is below.

```{r}
summary(cox1)

The forest plot for model 1 is shown below. Values closer to 1 indicate the hazard ratio for that covariate have less effect on the calculated hazard ratio.

```{r}
plot_model(cox1,
           dot.size = 2,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

Despite the high concordance value for model 1, the assumptions will
need to be evaluated before using the model to make predictions. If any
of the assumptions for the model are not true, the model will need to be
modified

#### Checking Proportional Hazards Assumption

@@

```
The new model has seven covariates with only one not meeting the proportional hazards assumption. Although sm5 does not have a p value > .05, the assumption overall holds for the new model and sm5 was not removed from the model because of its large impact on the hazard ratio shown in the forest plot. The new model's full summary is below.

@@

```
The plots of the martingale residuals vs the range of values for each covariate is shown above. For each covariate, there is no clear slope indicating the linearity assumption is true for the covariates in the model.

@@

@@

>>>>

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)
library(broom)

@@

plot_model(forest,
           dot.size = 1,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")

@@

### Survival Prediction

@@

Though CPH is typically used with health data, it does have applications
in other fields. This paper will study the remaining useful life (RUL)
of a NASA aircraft engine using data provided by NASA for the study on
propagation modeling [@saxena2008]. In particular, this paper will be
examining the data from the engine make two which has a training dataset
and a testing dataset.

**Data Structure**

Each engine in the NASA data has an unknown amount of wear,
manufacturing variation, and sensor noise which will impact the survival
time. In the training dataset, each engine is operating normally at the
beginning of the time series, will develop a fault, and experience
failure at some point. All engines will have failed in the training
dataset. The conditions the engines are subjected to are represented by
three operational setting fields and twenty-one sensor measurement
fields. In the testing dataset, none of the engines have experienced
failure [@saxena2008].

A column indicating status is needed for both the training and testing
datasets. The value 0 will be used to indicate if the machine is still
working and the value 1 will be used for the last cycle before the
engine fails. Both values will appear in the training dataset as all
engines in that dataset will eventually fail. The testing dataset will
only have the value of 0 as none of the machines have failed in that
dataset.

@@

@@

kable(head(test, n = 5), format = "markdown", caption = "Testing Data")

trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'

@@

kable(head(train, n = 5), format = "markdown", caption = 'Training Data')
```

**Data Visualization**

There are 260 engines in the training data. The histogram shows that the
survival times of the engines are normally distributed and the box plot
shows there six outliers in the dataset. The engines have a median
survival time of 199 cycles with a standard deviation of 46.8. The
maximum number of cycles an engine lasted was 378 while the minimum
number of cycles an engined lasted was 128.

```{r, echo = FALSE}
distributions <- train %>% 
  group_by(id) %>%
  summarise(time = max(time))

@@

colnames(summarymetrics) <- c("Metric", "Value")
summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
kable(summarymetrics, format = "markdown", padding = 0)

@@

To determine the intitial survival function, the Kaplan-Meier survival
curve was generated and is shown below. The graph verifies that the
median survival time is roughly 200 cycles and shows the maximum cycle.
After that cycle has been reached, the survival probability plummets to
0%.

@@

**Covariate Selection**

Based on the information provided by NASA on their data, there are
various sensors and operational setting fields that do not impact the
survival rate of an engine. To determine which covariates will impact
the survival probability, the below univariate test was run. The
covariates in which the p-value is $< 0.05$ will be considered for the
CPH models while the other covariates will be excluded.

@@

Based on the above test, the covariates that will be considered for the
CPH models are os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13,
sm14, sm19, sm20, and sm21 which is fourteen out of twenty-four possible
covariates.

**Generating Models**

Generation of all CPH models will be done using R, specifically the R
package `survival`. To generate the CPH model with properly formatted
survival data, a survival object is created by running
`Surv(time, event)` where `time` is the time field and `event` is the
event field. To generate the actual model, the survival object is then
applied to the function `coxph(survival object ~ x, data)` where `x`
represents one or multiple covariate(s) and `data` is the survival data
that is being modeled.[@R-base].

The table below provides the model number, the covariates used, and the
concordance index. The concordance index generates a value to determine
which model is the most accurate with the provided covariates. After
determining the most accurate model, the CPH assumptions will be tested.

@@

```{r}
concordance_results <- data.frame(c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", 
                                    "Model 6", "Model 7", "Model 8", "Model 9", "Model 10"), 
                   c("os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
                     "os1, os3, sm1, sm2, sm5", 
                     "os1, os3, sm1, sm5, sm6, sm7, sm13, sm14, sm19, sm21", 
                     "sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
                     "os1, os3, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm20", 
                     "os1, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm21",
                     "os1, sm13, sm14, sm19, sm20", 
                     "os1, os3, sm21",
                     "os1, os3, sm2, sm5, sm6, sm12, sm13, sm19, sm20, sm21",
                     "os3, sm1, sm2, sm5, sm6, sm11, sm12, sm13, sm19, sm20, sm21"),
                   c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6], cox5$concordance[6], cox6$concordance[6], cox7$concordance[6], cox8$concordance[6], cox9$concordance[6], cox10$concordance[6]))

colnames(concordance_results) <- c("Model", "Covariates", "Concordance Index")

kable(concordance_results, format = "markdown")

@@

Model 1 had the highest concordance index and used all of the covariates
that were determined to be significant to the survival of the engine.
The proceeding analysis will be done using model 1. The below tables
provide the summary values from the model.

```{r, echo = FALSE}
kable(tidy(cox1), format = "markdown")
g <- glance(cox1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")
```

The forest plot for model 1 is shown below. The covariates that are
closest to the value of one have a hazard rate that has less effect on
the calculated hazard ratio for the entire model.

```{r, echo = FALSE}
plot_model(cox1,
           dot.size = 2,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

After determining the model to use with the covariates, the CPH
assumptions need to be tested to confirm the model is a valid CPH model.
The model may have a high concordance index, but it may not be valid for
CPH to use to predict the survival probability. If any of the
assumptions for the model are not true, the model will need to be
modified.

@@

The new model has seven covariates with only one not meeting the
proportional hazards assumption. Although sm5 does not have a p value \>
.05, the assumption overall holds for the new model and sm5 was not
removed from the model because of its large impact on the hazard ratio
shown in the forest plot. The new model's full summary is below.

@@

The plots of the martingale residuals vs the range of values for each
covariate is shown above. For each covariate, there is no clear slope
indicating the linearity assumption is true for the covariates in the
model.

@@

### Adjusting the Model

@@

```{r}
ggsurvplot(survfit, data = test, fun = "pct", surv.median.line = "none")
```

```{r}
plot(survfit, mark.time=FALSE, xscale=365.24,
	xlab='Years', ylab='Survival')
lines(survfit[1], lwd=2, xscale=365.24)
```
